---
title: "Crossing interpretation for Morice (MORR)"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Crossing interpretation for Morice (MORR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(link)
```

Data generated by `data-raw/vignette_morr.R`. Sources: crossings from
fresh CSV, PSCIS assessments from BC Data Catalogue, override CSVs from
bcfishpass. No database required. The same steps run for any watershed
group code.

## The problem

Provincial crossing assessments produce a binary classification: BARRIER or
PASSABLE. This tells you nothing about how bad a barrier is or what habitat
it blocks.

```{r raw-crossings}
morr_raw <- readRDS(system.file(
  "testdata", "morr_crossings_raw.rds", package = "link"))

nrow(morr_raw)
table(morr_raw$crossing_source)
table(morr_raw$barrier_status)
```

MORR has `r nrow(morr_raw)` crossings — `r sum(morr_raw$crossing_source == "MODELLED CROSSINGS")` modelled, `r sum(morr_raw$crossing_source == "PSCIS")` with PSCIS field assessments. Most modelled crossings have no field data.

## Override corrections

bcfishpass maintains correction CSVs — tens of thousands of hand-reviewed
crossings accumulated across field seasons and imagery review.

```{r overrides}
morr_fixes <- readRDS(system.file(
  "testdata", "morr_modelled_fixes.rds", package = "link"))

nrow(morr_fixes)
table(morr_fixes$structure)
```

For MORR: `r nrow(morr_fixes)` modelled crossing fixes. Most reclassify
crossings as empty (no structure found on imagery) or open-bottom
structures (passable). These come from systematic review by multiple
people.

```{r overrides-sample}
head(morr_fixes[, c("modelled_crossing_id", "structure",
                     "reviewer_name", "source")], 8)
```

The link override pipeline loads, validates, and applies these corrections:

```{r overrides-pipeline, eval = FALSE}
conn <- lnk_db_conn()

# Load — validates CSV structure before writing to DB
lnk_override_load(conn,
  csv  = "user_modelled_crossing_fixes_morr.csv",
  to   = "working.morr_modelled_fixes",
  cols_id = "modelled_crossing_id",
  cols_required = c("structure"),
  cols_provenance = c("reviewer_name", "review_date", "source"))

# Validate — find orphans and duplicates
lnk_override_validate(conn,
  overrides = "working.morr_modelled_fixes",
  crossings = "working.morr_crossings")

# Apply — auto-detects which columns to update
lnk_override_apply(conn,
  crossings = "working.morr_crossings",
  overrides = "working.morr_modelled_fixes")
```

## PSCIS matching

PSCIS assessments have field measurements — outlet drop, culvert slope,
length, channel width. The crossings table has network position. Matching
links the measurements to the network so severity scoring can use real
data.

```{r pscis}
morr_pscis <- readRDS(system.file(
  "testdata", "morr_pscis.rds", package = "link"))
morr_xref <- readRDS(system.file(
  "testdata", "morr_xref.rds", package = "link"))

nrow(morr_pscis)
nrow(morr_xref)
```

The `r nrow(morr_xref)` xref corrections are hand-curated GPS error fixes
from field work. These take priority over spatial matching.

```{r pscis-match, eval = FALSE}
lnk_match_pscis(conn,
  crossings = "working.morr_crossings",
  pscis     = "working.morr_pscis",
  xref_csv  = "pscis_modelledcrossings_streams_xref_morr.csv",
  to        = "working.morr_matched")
```

## PSCIS barrier status overrides

```{r pscis-status}
morr_pscis_fixes <- readRDS(system.file(
  "testdata", "morr_pscis_status_fixes.rds", package = "link"))

if (nrow(morr_pscis_fixes) > 0) {
  morr_pscis_fixes[, c("stream_crossing_id",
                        "user_barrier_status", "notes")]
}
```

One MORR crossing has a barrier status override — confirmed PASSABLE by
imagery review.

## Severity scoring

Default thresholds classify crossings by biological impact:

```{r thresholds}
th <- lnk_thresholds()
```

| Severity | Criteria | Interpretation |
|----------|----------|----------------|
| High | outlet_drop >= `r th$high$outlet_drop`m OR slope*length >= `r th$high$slope_length` | Impassable at most flows |
| Moderate | outlet_drop >= `r th$moderate$outlet_drop`m OR slope*length >= `r th$moderate$slope_length` | Flow-dependent |
| Low | everything else | Likely passable |

```{r score, eval = FALSE}
lnk_score_severity(conn, "working.morr_crossings")
```

```{r severity-dist}
morr_dist <- readRDS(system.file(
  "testdata", "morr_severity_dist.rds", package = "link"))
morr_dist
```

```{r severity-detail}
morr_scored <- readRDS(system.file(
  "testdata", "morr_pscis_scored.rds", package = "link"))

# The reveal: crossings with PSCIS measurements scored by severity
scored_with_data <- morr_scored[!is.na(morr_scored$severity), ]
head(scored_with_data[order(-scored_with_data$outlet_drop),
  c("stream_crossing_id", "barrier_result_code",
    "outlet_drop", "culvert_slope",
    "severity")], 10)
```

The binary `barrier_result_code` treats these crossings identically. The
`severity` column reveals the gradient — from impassable drops to likely
passable culverts.

## Break source for fresh

`lnk_break_source()` produces the spec that plugs directly into
`frs_habitat()`. Severity labels translate to access labels via
`label_map`.

```{r break-source}
morr_spec <- readRDS(system.file(
  "testdata", "morr_break_source.rds", package = "link"))
str(morr_spec)
```

```{r break-source-use, eval = FALSE}
src <- lnk_break_source(conn, "working.morr_crossings")
fresh::frs_habitat(conn, "MORR", break_sources = list(src))
```

## Scale to any watershed group

The same steps work for any `watershed_group_code`. The override CSVs
have `watershed_group_code` columns for filtering. The crossings CSV from
fresh covers the entire province.

```{r province, eval = FALSE}
crossings <- read.csv(system.file("extdata", "crossings.csv",
                                   package = "fresh"))
wsgs <- unique(crossings$watershed_group_code)

for (wsg in wsgs) {
  # 1. Filter crossings to WSG
  # 2. Load and apply overrides (filtered to WSG)
  # 3. Fetch PSCIS from bcdata, match with lnk_match_pscis()
  # 4. lnk_score_severity()
  # 5. lnk_break_source() -> frs_habitat()
}
```
