---
title: "Crossing interpretation for Morice (MORR)"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Crossing interpretation for Morice (MORR)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(link)
```

Data generated by `data-raw/vignette_morr.R` against a PostgreSQL database
with bcfishpass and fwapg. The same steps run for any watershed group code.

## The problem

Provincial crossing assessments produce a binary classification: BARRIER or
PASSABLE. This tells you nothing about how bad a barrier is or what habitat
it blocks. Two crossings classified BARRIER can have completely different
biological impact — a 1.2m outlet drop versus a 0.1m drop with a steep
culvert.

link replaces binary classification with severity scoring based on actual
crossing measurements. The scored crossings feed into fresh for network
segmentation and habitat classification.

## Raw crossings

```{r raw-crossings}
morr_raw <- readRDS(system.file(
  "testdata", "morr_crossings_raw.rds", package = "link"))

nrow(morr_raw)

table(morr_raw$barrier_status)
```

MORR has `r nrow(morr_raw)` modelled crossings. Most are classified by the
model without field verification.

## Override corrections

bcfishpass maintains correction CSVs — tens of thousands of hand-reviewed
crossings accumulated across field seasons and imagery review. For MORR,
there are 947 modelled crossing fixes.

```{r overrides-data}
morr_fixes <- readRDS(system.file(
  "testdata", "morr_modelled_fixes.rds", package = "link"))

nrow(morr_fixes)
table(morr_fixes$structure)
head(morr_fixes[, c("modelled_crossing_id", "structure",
                     "reviewer_name", "source")], 8)
```

Most corrections reclassify crossings as NONE (no structure found on
imagery) or OBS (open bottom structure — passable). These come from
systematic imagery review by multiple reviewers.

```{r overrides-load, eval = FALSE}
conn <- lnk_db_conn()

# Load MORR modelled crossing fixes
lnk_override_load(conn,
  csv  = "data/overrides/user_modelled_crossing_fixes_morr.csv",
  to   = "working.morr_modelled_fixes",
  cols_id = "modelled_crossing_id",
  cols_required = c("structure"),
  cols_provenance = c("reviewer_name", "review_date", "source"))
```

Before applying, validate referential integrity:

```{r overrides-validate}
morr_validation <- readRDS(system.file(
  "testdata", "morr_overrides_validation.rds", package = "link"))

morr_validation$total_count
morr_validation$valid_count
length(morr_validation$orphans)
length(morr_validation$duplicates)
```

```{r overrides-validate-code, eval = FALSE}
lnk_override_validate(conn,
  overrides = "working.morr_modelled_fixes",
  crossings = "working.morr_crossings")
# Override validation: working.morr_modelled_fixes vs working.morr_crossings
#   Total overrides:  947
#   Valid (matched):  ...
#   Orphans:          ...  <-- crossings removed from model
#   Duplicates:       ...
```

Then apply:

```{r overrides-apply, eval = FALSE}
lnk_override_apply(conn,
  crossings = "working.morr_crossings",
  overrides = "working.morr_modelled_fixes",
  col_id = "modelled_crossing_id")
# Updated N of M crossings (structure)
```

```{r overrides-result}
morr_overrides <- readRDS(system.file(
  "testdata", "morr_crossings_overrides.rds", package = "link"))

table(morr_overrides$barrier_status)
```

The override corrections change the picture significantly — many crossings
that the model flagged as barriers are actually non-existent (NONE) or
passable structures (OBS).

## PSCIS matching

PSCIS assessments have field measurements — outlet drop, culvert slope,
length, channel width. Modelled crossings have network position. Matching
links the measurements to the network.

```{r pscis-data}
morr_pscis <- readRDS(system.file(
  "testdata", "morr_pscis.rds", package = "link"))
morr_xref <- readRDS(system.file(
  "testdata", "morr_xref.rds", package = "link"))

nrow(morr_pscis)
nrow(morr_xref)
```

The xref CSV contains `r nrow(morr_xref)` hand-curated matches for MORR —
GPS error corrections from field work. These take priority over spatial
matching.

```{r pscis-match, eval = FALSE}
lnk_match_pscis(conn,
  crossings = "working.morr_crossings",
  pscis     = "working.morr_pscis",
  xref_csv  = "data/overrides/pscis_modelledcrossings_streams_xref_morr.csv",
  to        = "working.morr_matched")
# Applied 8 known matches from xref CSV
# Matched N additional pairs spatially
```

```{r pscis-result}
morr_matched <- readRDS(system.file(
  "testdata", "morr_matched.rds", package = "link"))

nrow(morr_matched)
head(morr_matched[, c("source_a", "id_a", "source_b", "id_b", "distance_m")])
```

## PSCIS barrier status overrides

One MORR crossing has a barrier status override — crossing 197962
confirmed as PASSABLE (bridge) by imagery review.

```{r pscis-status}
morr_pscis_fixes <- readRDS(system.file(
  "testdata", "morr_pscis_status_fixes.rds", package = "link"))

morr_pscis_fixes[, c("stream_crossing_id", "user_barrier_status", "notes")]
```

```{r pscis-status-apply, eval = FALSE}
lnk_override_load(conn,
  csv   = "data/overrides/user_pscis_barrier_status_morr.csv",
  to    = "working.morr_pscis_status_fixes",
  cols_id = "stream_crossing_id")

lnk_override_apply(conn,
  crossings   = "working.morr_pscis",
  overrides   = "working.morr_pscis_status_fixes",
  col_id      = "stream_crossing_id",
  cols_update = c("barrier_result_code"),
  cols_provenance = NULL)
```

## Severity scoring

Default thresholds classify crossings by biological impact:

```{r thresholds}
th <- lnk_thresholds()
th
```

| Severity | Criteria | Interpretation |
|----------|----------|----------------|
| High | outlet_drop >= `r th$high$outlet_drop`m OR slope*length >= `r th$high$slope_length` | Impassable at most flows |
| Moderate | outlet_drop >= `r th$moderate$outlet_drop`m OR slope*length >= `r th$moderate$slope_length` | Flow-dependent |
| Low | everything else | Likely passable |

```{r score, eval = FALSE}
lnk_score_severity(conn, "working.morr_crossings")
```

```{r score-result}
morr_scored <- readRDS(system.file(
  "testdata", "morr_crossings_scored.rds", package = "link"))
morr_dist <- readRDS(system.file(
  "testdata", "morr_severity_dist.rds", package = "link"))

morr_dist
```

```{r score-comparison}
# The reveal: crossings with PSCIS measurements scored by severity
scored_with_pscis <- morr_scored[!is.na(morr_scored$outlet_drop), ]
if (nrow(scored_with_pscis) > 0) {
  head(scored_with_pscis[order(-scored_with_pscis$outlet_drop),
    c("modelled_crossing_id", "barrier_status",
      "outlet_drop", "culvert_slope", "culvert_length_m",
      "severity")], 10)
}
```

The binary `barrier_status` column treats these crossings identically.
The `severity` column reveals the gradient — from impassable drops to
likely passable culverts.

## Break source for fresh

`lnk_break_source()` produces the spec that plugs directly into
`frs_habitat()`. link's severity labels translate to fresh's access
labels via `label_map`.

```{r break-source}
morr_spec <- readRDS(system.file(
  "testdata", "morr_break_source.rds", package = "link"))

str(morr_spec)
```

```{r break-source-use, eval = FALSE}
# The handoff: link -> fresh
src <- lnk_break_source(conn, "working.morr_crossings")

# Feed to fresh for network segmentation and habitat classification
fresh::frs_habitat(conn, "MORR",
  break_sources = list(src))
```

## Scale to any watershed group

The same steps work for any `watershed_group_code`. To run the entire
province, loop over watershed groups:

```{r province, eval = FALSE}
wsgs <- DBI::dbGetQuery(conn,
  "SELECT DISTINCT watershed_group_code FROM bcfishpass.crossings")[[1]]

for (wsg in wsgs) {
  # 1. Extract crossings for this WSG
  # 2. Load and apply overrides (filtered to WSG)
  # 3. Match PSCIS (with xref CSV)
  # 4. Apply PSCIS barrier status overrides
  # 5. Score severity
  # 6. Produce break source -> frs_habitat()
}
```

The bcfishpass override CSVs already have `watershed_group_code` columns,
so filtering is straightforward. The thresholds and scoring logic are
identical across watershed groups — only the data changes.
