---
title: "Modelling spawning and rearing habitat using bcfishpass defaults"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Modelling spawning and rearing habitat using bcfishpass defaults}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 72
)
options(knitr.kable.NA = "—")
library(link)
```

This vignette documents link's reproduction of
[bcfishpass](https://github.com/smnorris/bcfishpass)'s linear habitat
classification — per-segment spawning and rearing booleans, conditioned
on natural accessibility — for the species present in five British
Columbia watershed groups (ADMS, BULK, BABL, ELKR, DEAD).

The bundled `"bcfishpass"` config encodes bcfishpass's per-species
rules and override CSVs as a self-contained input. Per-phase pipeline
detail lives in
[`research/bcfishpass_comparison.md`](https://github.com/NewGraphEnvironment/link/blob/main/research/bcfishpass_comparison.md).

## Prerequisites

Required:

- A PostgreSQL database with [fwapg](https://github.com/smnorris/fwapg)
  loaded — `wscode_ltree` / `localcode_ltree` columns on the FWA stream
  network plus the `fwa_upstream` / `fwa_downstream` /
  `fwa_watershedatmeasure` SQL functions the pipeline uses.

Recommended:

- [bcfishobs](https://github.com/smnorris/bcfishobs) populating
  `bcfishobs.observations` so the access model can override natural
  barriers per species.

Optional (for the comparison map at the bottom of this vignette only):

- A read-only tunnel to a bcfishpass reference database. Not needed to
  run link.

## Running the pipeline

Six `lnk_pipeline_*` phase calls driven by an `lnk_config()` bundle.
Each phase writes to a working schema in PostgreSQL; the final
`fresh.streams_habitat` table holds per-segment booleans for
`spawning`, `rearing`, `lake_rearing`, and `wetland_rearing`, per
species.

```{r entrypoint, eval = FALSE}
library(link)
library(DBI)

conn   <- lnk_db_conn()
cfg    <- lnk_config("bcfishpass")
schema <- "working_bulk"

lnk_pipeline_setup(   conn,                  schema, overwrite = TRUE)
lnk_pipeline_load(    conn, aoi = "BULK", cfg = cfg, schema)
lnk_pipeline_prepare( conn, aoi = "BULK", cfg = cfg, schema)
lnk_pipeline_break(   conn, aoi = "BULK", cfg = cfg, schema)
lnk_pipeline_classify(conn, aoi = "BULK", cfg = cfg, schema)
lnk_pipeline_connect( conn, aoi = "BULK", cfg = cfg, schema)
```

Run the same six phases per AOI to build a province-wide picture, or
plug them into a `targets` pipeline as
[`data-raw/_targets.R`](https://github.com/NewGraphEnvironment/link/blob/main/data-raw/_targets.R)
does for cross-WSG regression — Adams (ADMS), Bulkley (BULK), Babine
(BABL), Elk (ELKR), and Deadman (DEAD), binding per-WSG tibbles into
one rollup.

## The rollup

```{r load-rollup}
rollup <- readRDS(system.file("extdata", "vignette-data", "rollup.rds",
                               package = "link"))
```

Percent difference between link's classification and bcfishpass's
published `streams_habitat_linear` integer table (model + known habitat
combined) per species × watershed group. `diff_pct = (link_km −
bcfishpass_km) / bcfishpass_km × 100`. Negative = link under, positive
= link over. NA = species not in bcfishpass's published output.

```{r rollup-pivot, results = "asis"}
.pivot <- function(rollup, which_habitat) {
  x <- rollup[rollup$config == "bcfishpass" &
              rollup$habitat_type == which_habitat,
              c("species", "wsg", "diff_pct")]
  w <- stats::reshape(x, idvar = "species", timevar = "wsg",
                       direction = "wide", v.names = "diff_pct")
  names(w)[-1] <- sub("diff_pct\\.", "", names(w)[-1])
  cols <- intersect(c("species", "ADMS", "BULK", "BABL", "ELKR", "DEAD"),
    names(w))
  w <- w[order(w$species), cols]
  row.names(w) <- NULL
  w
}

knitr::kable(.pivot(rollup, "spawning"),
  digits = 1,
  caption = "Spawning parity (% diff vs bcfishpass)")

knitr::kable(.pivot(rollup, "rearing"),
  digits = 1,
  caption = "Rearing parity (% diff vs bcfishpass)")
```

All rows within ±5% of bcfishpass; median 1.5%. Residual deltas come
from segmentation-boundary rounding (per-segment attributes near rule
thresholds end up on different sides of the threshold) and a small
rearing-side stream-order bypass that's a documented future addition.

## Comparison map — Neexdzii Kwa (Upper Bulkley)

The watershed upstream of the Neexdzii Kwa / Wetzin Kwa (Bulkley /
Morice) confluence, built via `FWA_WatershedAtMeasure(360873822,
166030.4)`. Inside the BULK watershed group; the BULK rollup above
aggregates this area along with the rest of the Bulkley. The map
below visualizes the spawning/rearing slice for chinook (CH) — the
simplest cut available; per-species ha rollups, lake/wetland
decomposition, and other species are also in `fresh.streams_habitat`.
The link layer is visible by default; toggle on the bcfishpass
reference layer to compare.

```{r map-sub-ch, message = FALSE, warning = FALSE}
sub_ch      <- readRDS(system.file("extdata", "vignette-data",
                                    "sub_ch.rds", package = "link"))
sub_ch_bcfp <- readRDS(system.file("extdata", "vignette-data",
                                    "sub_ch_bcfp.rds", package = "link"))

if (requireNamespace("mapgl", quietly = TRUE)) {
  pal_values <- c("spawning only", "rearing only", "spawning + rearing")
  pal_colors <- c("#e31a1c", "#1f78b4", "#6a3d9a")

  mapgl::maplibre(
    bounds = sf::st_bbox(sub_ch),
    style = mapgl::carto_style("positron")
  ) |>
    mapgl::add_line_layer(
      id = "bcfishpass",
      source = sub_ch_bcfp,
      line_color = mapgl::match_expr("habitat",
        values = pal_values, stops = pal_colors, default = "#999999"),
      line_width = 3,
      line_opacity = 0.6,
      visibility = "none"
    ) |>
    mapgl::add_line_layer(
      id = "link",
      source = sub_ch,
      line_color = mapgl::match_expr("habitat",
        values = pal_values, stops = pal_colors, default = "#999999"),
      line_width = 2
    ) |>
    mapgl::add_legend(
      "Neexdzii Kwa (Upper Bulkley) · modelled chinook habitat",
      values = pal_values,
      colors = pal_colors,
      type = "categorical",
      position = "top-right"
    ) |>
    mapgl::add_layers_control(
      collapsible = TRUE,
      position    = "top-left"
    )
} else {
  message("Install `mapgl` (pak::pak('mapgl')) to render this map.")
}
```

## See also

- [`research/bcfishpass_comparison.md`](https://github.com/NewGraphEnvironment/link/blob/main/research/bcfishpass_comparison.md)
  — per-phase pipeline DAG, where breaks go, where classification
  comes from, known-habitat overlay, stream-order bypass, parity
  numbers
- [`inst/extdata/configs/bcfishpass/`](https://github.com/NewGraphEnvironment/link/tree/main/inst/extdata/configs/bcfishpass)
  — bundled config (rules YAML, dimensions CSV, per-species
  parameters, overrides)
- `?lnk_config`, `?lnk_pipeline_setup` and family — function reference
