---
title: "Bull trout and Arctic grayling: habitat and connectivity classification for the Parsnip River Watershed Group"
output:
  bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Bull trout and Arctic grayling: habitat and connectivity classification for the Parsnip River Watershed Group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150,
  message = FALSE,
  warning = FALSE
)
library(link)
```

For any stream in a watershed, fisheries managers want to know three things: can
a fish get there, is it good habitat, and what — if anything — blocks the way.
This vignette runs that analysis end to end for the Parsnip River Watershed
Group (`PARS`, ~5,600 km², north-eastern BC).

`link` models the entire stream network of a watershed group one segment at a
time, and for each species works out:

- **Access** — can the fish reach the segment, or is the route downstream
  likely too steep (gradient), or cut off by a barrier such as a dam,
  waterfall, or perched culvert?
- **Habitat** — if it is reachable, is the segment likely large enough
  (channel width) and its gradient gentle enough for the modelled species to
  use it as **spawning** habitat, **rearing** habitat, or simply passable water
  with neither?
- **Barriers** — the most significant obstacle downstream: a **dam**, a
  **known** barrier that has been assessed in the field (a road culvert, weir,
  etc.), a **modelled** crossing (predicted from road–stream intersections but
  not yet field-checked), or one that has since been **remediated** (fixed).

It also flags reaches that are **intermittent** (seasonally dry).
[bcfishpass](https://github.com/smnorris/bcfishpass) condenses that whole
per-segment verdict into one compact label it calls a `mapping_code` — for
example `SPAWN;DAM`, spawning habitat with a dam downstream. `link` reproduces
those same labels, and the maps below colour and weight every stream by them.

The vignette does two things. First, it shows `link`'s `bcfishpass`
configuration **reproducing** bcfishpass's per-segment classification for bull
trout (`BT`) — a parity check against the established tool. Second, it shows
`link` **extending** the same method to a species bcfishpass does not yet model
in the Peace: Arctic grayling (`GR`).

The Parsnip River Watershed Group sits between Prince George and Mackenzie, BC.
The Parsnip flows north into the southern arm of Williston Reservoir, joining
the Peace River system; from there the drainage runs Peace → Slave → Mackenzie,
ultimately discharging to the Arctic Ocean via the Mackenzie Delta. Of the
species in `link`'s `bcfishpass` configuration, bull trout (`BT`) is the only
one present, so the parity check below is a single-species comparison. Both
bull trout and Arctic grayling (`GR`) are cold-water species whose
distributions the model resolves through gradient, channel-width, and access
thresholds. Bull trout — provincially blue-listed and a COSEWIC species of
special concern in the Western Arctic population — spawn as adfluvial migrants
in cold, low-gradient tributaries such as the Misinchinka and Anzac. Arctic
grayling, at the southern edge of their range in the Williston watershed, hold
to cooler, larger (fourth-order and up) clear-water reaches and spawn over fine
gravels.

`link` is layered on the canonical
[`fwapg`](https://github.com/smnorris/fwapg) /
`bcfishpass` / [`bcfishobs`](https://github.com/smnorris/bcfishobs) stack from
Simon Norris (Hillcrest Geographics). What `link` adds is a configuration-driven
re-expression of the same modelling: we can experiment with different
configurations for species bcfishpass already models, and extend to species it
does not — here, Arctic grayling — while staying byte-checkable against the
upstream reference.

## Modelling parameters

A `mapping_code` is a per-segment, per-species label that bcfishpass computes
over the BC Freshwater Atlas (FWA) stream network — it is not part of the FWA
itself. Producing it means first re-cutting the FWA streams into shorter
segments wherever a fish's prospects change — at gradient transitions, falls,
dams, modelled and assessed crossings, and habitat thresholds — then giving each
resulting segment its access, spawning, and rearing classification. `link`
reproduces that segmentation and those classifications. Access is gated by a
per-species **maximum** gradient; spawning and rearing are then gated by their
own **maximum** gradients and a minimum channel width. The values in force for
this run are below.

```{r params, echo = FALSE}
params <- data.frame(
  species = c("BT", "GR"),
  configuration = c("bcfishpass (parity)", "default (link extension)"),
  access_grad_max = c(0.25, 0.15),
  spawn_grad_max = c(0.0549, 0.0249),
  rear_grad_max = c(0.1049, 0.0349),
  spawn_cw_min_m = c(2, 4),
  rear_cw_min_m = c(1.5, 1.5),
  stringsAsFactors = FALSE
)
knitr::kable(
  params, row.names = FALSE,
  col.names = c("species", "configuration", "access grad max",
                "spawn grad max", "rear grad max", "spawn CW min (m)",
                "rear CW min (m)"),
  caption = paste0(
    "Access, spawning, and rearing gradient ceilings and minimum channel ",
    "widths for the two species mapped in this vignette. Bull trout is ",
    "modelled by both bcfishpass and link's bcfishpass config (the parity ",
    "case); Arctic grayling is modelled only by link's default config (the ",
    "extension). Grayling's lower access ceiling (0.15 vs 0.25) and narrower ",
    "spawning/rearing gradient windows are what give it the smaller modelled ",
    "network seen below."
  )
)
```

Each segment's `mapping_code` is a compact token of the form
`<use>;<barrier-status>`, optionally suffixed `;INTERMITTENT` for intermittent
streams. The first field is the highest-value habitat use modelled for that
segment — `SPAWN`, `REAR`, or `ACCESS` (reachable, but no modelled spawning or
rearing habitat). The second records the most significant barrier *downstream*
of the segment: `NONE` (none known), `MODELLED` (a modelled potential barrier),
`ASSESSED` (a field-assessed known barrier), `DAM`, or `REMEDIATED` (a barrier
since fixed). Stream colour is keyed on barrier status alone — a purple segment
sits below a dam, a red one below a field-assessed barrier — regardless of
habitat use, while line width encodes the habitat use itself: spawning reaches
draw thickest, rearing medium, access-only thinnest. Intermittent reaches
(`;INTERMITTENT`) draw dashed. Colours and widths are read straight from the
bcfishpass symbology registry, so the maps match a bcfishpass QGIS project
exactly.

## Cached inputs

The model run and the bcfishpass comparison both require a populated
PostgreSQL/PostGIS database and a bcfishpass snapshot — neither of which
exists on the documentation-build CI. So they run **once, locally**, in
`data-raw/wsg_vignette_data.R` (generic — set `aoi` and re-run for any
watershed group), which caches its outputs to `inst/vignette-data/`. This
vignette only *loads* those artifacts; it never touches a database at build
time. Direct downloads from the repo (open in QGIS or any GDAL-aware tool):

- [`pars.gpkg`](https://github.com/NewGraphEnvironment/link/raw/main/inst/vignette-data/pars.gpkg)
  — vectors: `aoi` (WSG boundary), `streams` (per-segment `mapping_code_bt`
  from the bcfishpass config + `mapping_code_gr` from the default config),
  `waterbodies` (lakes + rivers + manmade), `named_streams`, and basemapping
  context layers `reserves`, `parks`, `roads`, `railways`
- [`pars_parity.rds`](https://github.com/NewGraphEnvironment/link/raw/main/inst/vignette-data/pars_parity.rds)
  — tunnel-free per-species `mapping_code` parity tibble

```{r load, include = FALSE}
library(sf)

gpkg   <- system.file("vignette-data/pars.gpkg",   package = "link", mustWork = TRUE)
parity <- readRDS(system.file("vignette-data/pars_parity.rds",
                              package = "link", mustWork = TRUE))

present <- sf::st_layers(gpkg)$name
lyr <- function(name) {
  if (name %in% present) sf::st_read(gpkg, layer = name, quiet = TRUE) else NULL
}

aoi           <- lyr("aoi")
streams       <- lyr("streams")
waterbodies   <- lyr("waterbodies")
named_streams <- lyr("named_streams")
reserves      <- lyr("reserves")
parks         <- lyr("parks")
roads         <- lyr("roads")
railways      <- lyr("railways")

# Context layers (reserves/parks/roads/railways) are pulled by intersection
# with the WSG boundary, so whole features can spill past it. Clip them to the
# boundary shape so nothing renders outside the basin.
clip_aoi <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(x)
  out <- suppressWarnings(sf::st_intersection(x, sf::st_union(sf::st_geometry(aoi))))
  if (nrow(out) == 0L) NULL else out
}
roads    <- clip_aoi(roads)
railways <- clip_aoi(railways)
reserves <- clip_aoi(reserves)
parks    <- clip_aoi(parks)
```

The model run itself — shown here for reference, not executed at build time —
is one call per configuration. The Peace study-area run that produced the
state this vignette reads modelled the drainage **most-downstream-watershed-group
first**, so a segment's downstream-dam (`;DAM`) tokens, which can live in an
adjacent watershed group, resolve correctly. A standalone single-group re-run
would diverge on exactly those cross-group segments, so the data-gen script
reads the persisted study-area state rather than recomputing it.

```{r model-run, eval = FALSE}
conn <- lnk_db_conn()

cfg_bcfp <- lnk_config("bcfishpass")          # persists to schema `fresh`
loaded   <- lnk_load_overrides(cfg_bcfp)
lnk_pipeline_run(conn, aoi = "PARS", cfg = cfg_bcfp, loaded = loaded,
                 mapping_code = TRUE)

cfg_default <- lnk_config("default")          # persists to `fresh_default`
cfg_default$pipeline$schema <- "fresh_default"
loaded_d <- lnk_load_overrides(cfg_default)
lnk_pipeline_run(conn, aoi = "PARS", cfg = cfg_default, loaded = loaded_d,
                 mapping_code = TRUE)
```

## Reproducing bcfishpass (parity)

`lnk_compare_mapping_code()` compares `link`'s per-segment `mapping_code`
against the local bcfishpass snapshot, segment by segment, with no database
tunnel required. The comparison restricts itself to species that are actually
active in the watershed group — which, for the reasons above, is bull trout
alone in `PARS`.

```{r parity-table, echo = FALSE}
ptab <- parity[, c("wsg", "species", "total_segs", "match_pct",
                   "n_diffs", "top_pattern", "top_pattern_count")]
names(ptab) <- c("WSG", "species", "segments", "match %", "n diffs",
                 "top diff pattern (link | bcfishpass)", "count")
knitr::kable(
  ptab, row.names = FALSE,
  caption = paste0(
    "Per-segment mapping_code parity for bull trout in PARS, link's ",
    "bcfishpass config vs the local bcfishpass snapshot. The top diff pattern ",
    "column shows the most common (link | reference) disagreement, not a ",
    "literal mapping_code value."
  )
)
```

```{r parity-pct, echo = FALSE, results = "asis"}
cat(sprintf(paste0("link reproduces **%.2f%%** of bcfishpass's per-segment ",
                   "bull-trout `mapping_code` across %s segments, with %s ",
                   "disagreements. That is consistent with the 99.66%% ",
                   "study-area median established for the Peace.\n"),
            parity$match_pct[1],
            format(parity$total_segs[1], big.mark = ","),
            format(parity$n_diffs[1], big.mark = ",")))
```

The remaining disagreements concentrate on intermittent reaches downstream of
dams — segments where the `;INTERMITTENT` and `;DAM` qualifiers interact, and
where cross-watershed-group ordering is most sensitive.

```{r symbology, include = FALSE}
reg <- gq::gq_reg_main()
cls <- gq::gq_tmap_classes(reg$layers$streams_salmon)

# bcfishpass line widths come from the same registry — habitat use drives line
# weight (SPAWN thick, REAR medium, ACCESS thin), barrier status drives colour.
# Stored raw here and scaled per map below (full-WSG views need thinner lines
# than the zoomed detail panels). Unmatched tokens fall to the thinnest weight.
width_for <- function(tokens) {
  w <- cls$widths[tokens]
  w[is.na(w)] <- min(cls$widths, na.rm = TRUE)
  unname(w)
}

# Legend labels are decoded from the mapping_code token rather than taken from
# cls$labels: the upstream QGIS project mislabels the ACCESS habitat token as
# "No known barriers" (colliding with token2's own status wording, producing
# "No known barriers; known barrier"). Fix-at-source tracked in the bcfishpass
# fork and gq registry; decoded here so the legend reads correctly meanwhile.
.tok1 <- c(SPAWN = "Spawning", REAR = "Rearing", ACCESS = "Accessible")
.tok2 <- c(NONE = "no known barriers", MODELLED = "potential barrier",
           ASSESSED = "known barrier", DAM = "dam", REMEDIATED = "remediated")
label_for <- function(tokens) {
  vapply(strsplit(tokens, ";", fixed = TRUE), function(p) {
    parts <- c(.tok1[p[1]], if (length(p) >= 2) .tok2[p[2]],
               if (length(p) >= 3 && p[3] == "INTERMITTENT") "intermittent")
    paste(parts[!is.na(parts)], collapse = "; ")
  }, character(1))
}

# Context layers shared by the two full-WSG maps, drawn UNDER the
# mapping_code streams: waterbodies, resource roads, railways, then the
# land-use polygons (parks, reserves) on top of hydro/transport so they read
# where they overlap — the flooded layering. Reserves are tiny at WSG scale,
# so each gets a centroid diamond marker.
draw_context_full <- function() {
  if (!is.null(waterbodies)) {
    plot(sf::st_geometry(waterbodies), col = "#a3cdb966", border = "#2171B5",
         lwd = 0.3, add = TRUE)
  }
  if (!is.null(roads)) {
    plot(sf::st_geometry(roads), col = "#88888888", lwd = 0.25, add = TRUE)
  }
  if (!is.null(railways)) {
    plot(sf::st_geometry(railways), col = "black", lwd = 0.8, lty = "dashed",
         add = TRUE)
  }
  if (!is.null(parks)) {
    plot(sf::st_geometry(parks), col = "#639b5f55", border = "#33a02c",
         lwd = 0.7, add = TRUE)
  }
}

# Reserves draw OVER the mapping_code streams (called after the stream plot),
# so the polygon + centroid diamond read where they overlap the network.
draw_reserves <- function(x, fill = "#b2b2b288", cex = 0.9) {
  if (is.null(x) || nrow(x) == 0L) return(invisible())
  plot(sf::st_geometry(x), col = fill, border = "#232323", lwd = 0.6,
       add = TRUE)
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(x),
                                          of_largest_polygon = TRUE))
  plot(pts, pch = 23, bg = "#222222", col = "white", cex = cex, add = TRUE)
}

# Reserve english_name labels with a thin white halo, drawn last with
# xpd = TRUE so an edge label isn't cropped at the plot region.
label_reserves <- function(x, cex = 0.45) {
  if (is.null(x) || !"english_name" %in% names(x) || nrow(x) == 0L) {
    return(invisible())
  }
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(x),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  ofs_y  <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) {
    text(coords[, 1] + dx * halo_x, ly, labels = x$english_name,
         cex = cex, col = "white", font = 2, xpd = TRUE)
  }
  for (dy in c(-1, 1)) {
    text(coords[, 1], ly + dy * halo_y, labels = x$english_name,
         cex = cex, col = "white", font = 2, xpd = TRUE)
  }
  text(coords[, 1], ly, labels = x$english_name, cex = cex, col = "#111111",
       font = 2, xpd = TRUE)
}

# mapping_code legend drawn outside the plotting region (into the reserved
# right margin) so it never overlaps the network.
legend_mapping_code <- function(tokens) {
  present <- names(cls$values) %in% unique(tokens)
  toks <- names(cls$values)[present]
  key_lwd <- cls$widths[present] * 2
  # base R scales dash length by line width, so a thick "dashed" key fits only
  # ~1 dash in the short legend segment while thin keys show many. Give thicker
  # intermittent keys a finer pattern so all read as dashes at the same seg.len.
  is_int <- grepl(";INTERMITTENT", toks)
  key_lty <- ifelse(!is_int, "solid",
                    ifelse(key_lwd > 2.5, "22",
                           ifelse(key_lwd > 1.5, "33", "44")))
  legend("topright", inset = c(-0.34, 0), xpd = TRUE, title = "Legend",
         legend = label_for(toks), col = cls$values[present],
         lwd = key_lwd, lty = key_lty,
         seg.len = 3, cex = 0.55, bty = "n")
}
```

```{r map-bt, echo = FALSE, fig.cap = "Bull-trout per-segment mapping_code across the Parsnip River Watershed Group, link's bcfishpass configuration. Stream colours come straight from the bcfishpass symbology registry, so they match a bcfishpass QGIS project exactly. Context: lakes/rivers/manmade waterbodies (light blue), provincial parks (green), First Nations reserves (grey polygon + black diamond + label), resource roads (grey), railways (black dashed); the heavy black line is the watershed-group boundary."}
bt <- streams[!is.na(streams$mapping_code_bt) & nzchar(streams$mapping_code_bt), ]
bt$col <- cls$values[bt$mapping_code_bt]
bt$col[is.na(bt$col)] <- "#999999"
bt$w <- width_for(bt$mapping_code_bt)
bt$lty <- ifelse(grepl(";INTERMITTENT", bt$mapping_code_bt), "dashed", "solid")

op <- par(mar = c(2, 1, 4, 8))
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, axes = FALSE,
     main = "PARS — bull trout mapping_code\n(link bcfishpass config)")
draw_context_full()
plot(sf::st_geometry(bt), col = bt$col, lwd = bt$w * 0.6, lty = bt$lty, add = TRUE)
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, add = TRUE)
draw_reserves(reserves)
label_reserves(reserves)
legend_mapping_code(bt$mapping_code_bt)
par(op)
```

## Arctic grayling — a link extension

bcfishpass does not yet model Arctic grayling, so there is nothing to compare
against; this is net-new output. `link`'s `default` configuration carries `GR`
in its species dimensions, and the same six-phase pipeline that produced the
bull-trout parity above produces a per-segment `mapping_code` for grayling.
The map below is rendered with the **same** bcfishpass symbology registry — the
token vocabulary (`ACCESS`/`SPAWN`/`REAR` × `NONE`/`MODELLED`/`ASSESSED`/`DAM`/…)
is species-agnostic, so one colour lookup styles every species consistently.

```{r map-gr, echo = FALSE, fig.cap = "Arctic grayling per-segment mapping_code across the Parsnip River Watershed Group, link's default configuration — a species bcfishpass does not yet model. Same symbology registry and context layers as the bull-trout map, so the two are directly comparable. The grayling network is smaller than bull trout's (19,233 vs 31,932 classified segments), but every grayling segment is net-new output relative to bcfishpass, and 1,764 of them carry no bull-trout classification at all."}
gr <- streams[!is.na(streams$mapping_code_gr) & nzchar(streams$mapping_code_gr), ]
gr$col <- cls$values[gr$mapping_code_gr]
gr$col[is.na(gr$col)] <- "#999999"
gr$w <- width_for(gr$mapping_code_gr)
gr$lty <- ifelse(grepl(";INTERMITTENT", gr$mapping_code_gr), "dashed", "solid")

op <- par(mar = c(2, 1, 4, 8))
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, axes = FALSE,
     main = "PARS — Arctic grayling mapping_code\n(link default config)")
draw_context_full()
plot(sf::st_geometry(gr), col = gr$col, lwd = gr$w * 0.6, lty = gr$lty, add = TRUE)
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, add = TRUE)
draw_reserves(reserves)
label_reserves(reserves)
legend_mapping_code(gr$mapping_code_gr)
par(op)
```

## Maps — detail comparison

The full-watershed views compress a lot of network. Cropping to a sub-reach
puts bull trout and grayling side by side at full resolution. Grayling's
modelled network is the smaller of the two overall, but 1,764 segments carry a
grayling classification with no bull-trout classification — the reaches where
the extension is doing genuinely new work.

```{r map-detail, echo = FALSE, fig.width = 7, fig.height = 4.5, fig.cap = "South-east corner of the Parsnip River Watershed Group at full resolution — the headwaters near the continental divide: bull trout (left, link bcfishpass config) and Arctic grayling (right, link default config), same extent, same symbology. Grey background streams are the full modelled network, so the coloured overlay shows where each species' classification reaches. Context: waterbodies (light blue), parks (green), reserves (grey + diamond), roads (grey), railways (black dashed), named streams (italic blue labels)."}
e <- sf::st_bbox(aoi)
inset_bbox <- sf::st_bbox(c(
  xmin = unname(e["xmin"] + (e["xmax"] - e["xmin"]) * 0.55),
  ymin = unname(e["ymin"]),
  xmax = unname(e["xmax"]),
  ymax = unname(e["ymin"] + (e["ymax"] - e["ymin"]) * 0.45)
), crs = sf::st_crs(aoi))

crop_sf <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(NULL)
  out <- suppressWarnings(sf::st_crop(x, inset_bbox))
  if (nrow(out) == 0L) NULL else out
}

frame <- sf::st_as_sfc(inset_bbox)
all_d   <- crop_sf(streams)
bt_d    <- crop_sf(bt)
gr_d    <- crop_sf(gr)
wb_d    <- crop_sf(waterbodies)
aoi_d   <- crop_sf(aoi)
roads_d <- crop_sf(roads)
rail_d  <- crop_sf(railways)
res_d   <- crop_sf(reserves)
parks_d <- crop_sf(parks)
ns_d    <- crop_sf(named_streams)

panel <- function(seg, title) {
  plot(sf::st_geometry(frame), col = NA, border = NA, axes = FALSE,
       main = title)
  if (!is.null(wb_d)) {
    plot(sf::st_geometry(wb_d), col = "#a3cdb988", border = "#2171B5",
         lwd = 0.4, add = TRUE)
  }
  if (!is.null(all_d)) {
    plot(sf::st_geometry(all_d), col = "#cccccc", lwd = 0.4, add = TRUE)
  }
  if (!is.null(roads_d)) {
    plot(sf::st_geometry(roads_d), col = "#666666aa", lwd = 0.4, add = TRUE)
  }
  if (!is.null(rail_d)) {
    plot(sf::st_geometry(rail_d), col = "black", lwd = 1.0, lty = "dashed",
         add = TRUE)
  }
  if (!is.null(parks_d)) {
    plot(sf::st_geometry(parks_d), col = "#639b5f55", border = "#33a02c",
         lwd = 0.8, add = TRUE)
  }
  if (!is.null(seg) && nrow(seg) > 0L) {
    plot(sf::st_geometry(seg), col = seg$col, lwd = seg$w, lty = seg$lty,
         add = TRUE)
  }
  draw_reserves(res_d, fill = "#b2b2b2aa", cex = 1.0)
  if (!is.null(aoi_d)) {
    plot(sf::st_geometry(aoi_d), border = "black", lwd = 1.5, add = TRUE)
  }
  if (!is.null(ns_d)) {
    dedup <- ns_d[!duplicated(ns_d$gnis_name), ]
    npts <- suppressWarnings(sf::st_centroid(sf::st_geometry(dedup)))
    text(sf::st_coordinates(npts), labels = dedup$gnis_name, cex = 0.45,
         font = 3, col = "#0d3a6c")
  }
}

op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
panel(bt_d, "Bull trout")
panel(gr_d, "Arctic grayling")
par(op)
```
