---
title: "Per-segment habitat mapping_code — Parsnip River Watershed Group"
output:
  bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Per-segment habitat mapping_code — Parsnip River Watershed Group}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi = 150,
  message = FALSE,
  warning = FALSE
)
library(link)
```

This vignette walks a per-segment habitat `mapping_code` analysis through the
Parsnip River Watershed Group (`PARS`, ~5,600 km², north-eastern BC) end to
end. It does two things. First, it shows `link`'s `bcfishpass` configuration
**reproducing** the per-segment `mapping_code` that
[bcfishpass](https://github.com/smnorris/bcfishpass) produces — the parity
check that earns trust in the rest of the analysis. Second, it shows `link`
**extending** that methodology to a species bcfishpass does not model in the
Peace: Arctic grayling (`GR`).

The Parsnip River Watershed Group sits between Prince George and Mackenzie, BC.
The Parsnip flows north into the southern arm of Williston Reservoir, joining
the Peace River system; from there the drainage runs Peace → Slave → Mackenzie,
ultimately discharging to the Arctic Ocean via the Mackenzie Delta. It lies
**above the W.A.C. Bennett Dam**, which severs the upper Peace from the Pacific,
so the Pacific salmon and steelhead bcfishpass models elsewhere do not occur
here — of the species in `link`'s `bcfishpass` configuration, bull trout
(`BT`) is the only one present, which is why the parity check below is a
single-species comparison. Arctic grayling, a defining sportfish of the upper
Peace that bcfishpass does not model, is the extension.

`link` is layered on the canonical
[`fwapg`](https://github.com/smnorris/fwapg) /
`bcfishpass` / [`bcfishobs`](https://github.com/smnorris/bcfishobs) stack from
Simon Norris (Hillcrest Geographics). That stack is the foundation; `link` does
not replace it. What `link` adds is a configuration-driven re-expression of the
same modelling so it can be pointed at new species and new study areas — here,
the **Fish Passage Peace 2025** program — while staying byte-checkable against
the upstream reference.

## Modelling parameters

Per-segment `mapping_code` is assembled from access, spawning, and rearing
classifications. Access is gated by a per-species **maximum** gradient; spawning
and rearing are then gated by their own **maximum** gradients and a minimum
channel width. The values in force for this run are below.

```{r params, echo = FALSE}
params <- data.frame(
  species = c("BT", "GR"),
  configuration = c("bcfishpass (parity)", "default (link extension)"),
  access_grad_max = c(0.25, 0.15),
  spawn_grad_max = c(0.0549, 0.0249),
  rear_grad_max = c(0.1049, 0.0349),
  spawn_cw_min_m = c(2, 4),
  rear_cw_min_m = c(1.5, 1.5),
  modelled_by_bcfishpass = c("yes", "no"),
  stringsAsFactors = FALSE
)
knitr::kable(
  params, row.names = FALSE,
  col.names = c("species", "configuration", "access grad max",
                "spawn grad max", "rear grad max", "spawn CW min (m)",
                "rear CW min (m)", "modelled by bcfishpass"),
  caption = paste0(
    "Access, spawning, and rearing gradient ceilings and minimum channel ",
    "widths for the two species mapped in this vignette. Bull trout is ",
    "modelled by both bcfishpass and link's bcfishpass config (the parity ",
    "case); Arctic grayling is modelled only by link's default config (the ",
    "extension). Grayling's lower access ceiling (0.15 vs 0.25) and narrower ",
    "spawning/rearing gradient windows are what give it the smaller modelled ",
    "network seen below."
  )
)
```

## Cached inputs

The model run and the bcfishpass comparison both require a populated
PostgreSQL/PostGIS database and a bcfishpass snapshot — neither of which
exists on the documentation-build CI. So they run **once, locally**, in
`data-raw/wsg_vignette_data.R` (generic — set `aoi` and re-run for any
watershed group), which caches its outputs to `inst/vignette-data/`. This
vignette only *loads* those artifacts; it never touches a database at build
time. Direct downloads from the repo (open in QGIS or any GDAL-aware tool):

- [`pars.gpkg`](https://github.com/NewGraphEnvironment/link/raw/main/inst/vignette-data/pars.gpkg)
  — vectors: `aoi` (WSG boundary), `streams` (per-segment `mapping_code_bt`
  from the bcfishpass config + `mapping_code_gr` from the default config),
  `waterbodies` (lakes + rivers + manmade), `named_streams`, and basemapping
  context layers `reserves`, `parks`, `roads`, `railways`
- [`pars_parity.rds`](https://github.com/NewGraphEnvironment/link/raw/main/inst/vignette-data/pars_parity.rds)
  — tunnel-free per-species `mapping_code` parity tibble

```{r load, include = FALSE}
library(sf)

gpkg   <- system.file("vignette-data/pars.gpkg",   package = "link", mustWork = TRUE)
parity <- readRDS(system.file("vignette-data/pars_parity.rds",
                              package = "link", mustWork = TRUE))

present <- sf::st_layers(gpkg)$name
lyr <- function(name) {
  if (name %in% present) sf::st_read(gpkg, layer = name, quiet = TRUE) else NULL
}

aoi           <- lyr("aoi")
streams       <- lyr("streams")
waterbodies   <- lyr("waterbodies")
named_streams <- lyr("named_streams")
reserves      <- lyr("reserves")
parks         <- lyr("parks")
roads         <- lyr("roads")
railways      <- lyr("railways")

# Context layers (reserves/parks/roads/railways) are pulled by intersection
# with the WSG boundary, so whole features can spill past it. Clip them to the
# boundary shape so nothing renders outside the basin.
clip_aoi <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(x)
  out <- suppressWarnings(sf::st_intersection(x, sf::st_union(sf::st_geometry(aoi))))
  if (nrow(out) == 0L) NULL else out
}
roads    <- clip_aoi(roads)
railways <- clip_aoi(railways)
reserves <- clip_aoi(reserves)
parks    <- clip_aoi(parks)
```

The model run itself — shown here for reference, not executed at build time —
is one call per configuration. The Peace study-area run that produced the
state this vignette reads modelled the drainage **most-downstream-watershed-group
first**, so a segment's downstream-dam (`;DAM`) tokens, which can live in an
adjacent watershed group, resolve correctly. A standalone single-group re-run
would diverge on exactly those cross-group segments, so the data-gen script
reads the persisted study-area state rather than recomputing it.

```{r model-run, eval = FALSE}
conn <- lnk_db_conn()

cfg_bcfp <- lnk_config("bcfishpass")          # persists to schema `fresh`
loaded   <- lnk_load_overrides(cfg_bcfp)
lnk_pipeline_run(conn, aoi = "PARS", cfg = cfg_bcfp, loaded = loaded,
                 mapping_code = TRUE)

cfg_default <- lnk_config("default")          # persists to `fresh_default`
cfg_default$pipeline$schema <- "fresh_default"
loaded_d <- lnk_load_overrides(cfg_default)
lnk_pipeline_run(conn, aoi = "PARS", cfg = cfg_default, loaded = loaded_d,
                 mapping_code = TRUE)
```

## Reproducing bcfishpass (parity)

`lnk_compare_mapping_code()` compares `link`'s per-segment `mapping_code`
against the local bcfishpass snapshot, segment by segment, with no database
tunnel required. The comparison restricts itself to species that are actually
active in the watershed group — which, for the reasons above, is bull trout
alone in `PARS`.

```{r parity-table, echo = FALSE}
ptab <- parity[, c("wsg", "species", "total_segs", "match_pct",
                   "n_diffs", "top_pattern", "top_pattern_count")]
names(ptab) <- c("WSG", "species", "segments", "match %", "n diffs",
                 "top diff pattern (link | bcfishpass)", "count")
knitr::kable(
  ptab, row.names = FALSE,
  caption = paste0(
    "Per-segment mapping_code parity for bull trout in PARS, link's ",
    "bcfishpass config vs the local bcfishpass snapshot. The top diff pattern ",
    "column shows the most common (link | reference) disagreement, not a ",
    "literal mapping_code value."
  )
)
```

```{r parity-pct, echo = FALSE, results = "asis"}
cat(sprintf(paste0("link reproduces **%.2f%%** of bcfishpass's per-segment ",
                   "bull-trout `mapping_code` across %s segments, with %s ",
                   "disagreements. That is consistent with the 99.66%% ",
                   "study-area median established for the Peace.\n"),
            parity$match_pct[1],
            format(parity$total_segs[1], big.mark = ","),
            format(parity$n_diffs[1], big.mark = ",")))
```

The remaining disagreements concentrate on intermittent reaches downstream of
dams — segments where the `;INTERMITTENT` and `;DAM` qualifiers interact, and
where cross-watershed-group ordering is most sensitive.

```{r symbology, include = FALSE}
reg <- gq::gq_reg_main()
cls <- gq::gq_tmap_classes(reg$layers$streams_salmon)

# Context layers shared by the two full-WSG maps, drawn UNDER the
# mapping_code streams: waterbodies, resource roads, railways, then the
# land-use polygons (parks, reserves) on top of hydro/transport so they read
# where they overlap — the flooded layering. Reserves are tiny at WSG scale,
# so each gets a centroid diamond marker.
draw_context_full <- function() {
  if (!is.null(waterbodies)) {
    plot(sf::st_geometry(waterbodies), col = "#a3cdb966", border = "#2171B5",
         lwd = 0.3, add = TRUE)
  }
  if (!is.null(roads)) {
    plot(sf::st_geometry(roads), col = "#88888888", lwd = 0.25, add = TRUE)
  }
  if (!is.null(railways)) {
    plot(sf::st_geometry(railways), col = "black", lwd = 0.8, lty = "dashed",
         add = TRUE)
  }
  if (!is.null(parks)) {
    plot(sf::st_geometry(parks), col = "#639b5f55", border = "#33a02c",
         lwd = 0.7, add = TRUE)
  }
  if (!is.null(reserves)) {
    plot(sf::st_geometry(reserves), col = "#b2b2b288", border = "#232323",
         lwd = 0.6, add = TRUE)
    pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(reserves),
                                            of_largest_polygon = TRUE))
    plot(pts, pch = 23, bg = "#222222", col = "white", cex = 0.9, add = TRUE)
  }
}

# Reserve english_name labels with a thin white halo, drawn last with
# xpd = TRUE so an edge label isn't cropped at the plot region.
label_reserves <- function(x, cex = 0.45) {
  if (is.null(x) || !"english_name" %in% names(x) || nrow(x) == 0L) {
    return(invisible())
  }
  pts <- suppressWarnings(sf::st_centroid(sf::st_geometry(x),
                                          of_largest_polygon = TRUE))
  coords <- sf::st_coordinates(pts)
  ofs_y  <- par("cxy")[2] * 0.45 * 0.9
  halo_x <- par("cxy")[1] * 0.45 * 0.05
  halo_y <- par("cxy")[2] * 0.45 * 0.05
  ly <- coords[, 2] - ofs_y
  for (dx in c(-1, 1)) {
    text(coords[, 1] + dx * halo_x, ly, labels = x$english_name,
         cex = cex, col = "white", font = 2, xpd = TRUE)
  }
  for (dy in c(-1, 1)) {
    text(coords[, 1], ly + dy * halo_y, labels = x$english_name,
         cex = cex, col = "white", font = 2, xpd = TRUE)
  }
  text(coords[, 1], ly, labels = x$english_name, cex = cex, col = "#111111",
       font = 2, xpd = TRUE)
}

# mapping_code legend drawn outside the plotting region (into the reserved
# right margin) so it never overlaps the network.
legend_mapping_code <- function(tokens) {
  present <- names(cls$values) %in% unique(tokens)
  legend("topright", inset = c(-0.34, 0), xpd = TRUE, title = "mapping_code",
         legend = cls$labels[present], col = cls$values[present],
         lwd = 2, cex = 0.55, bty = "n")
}
```

```{r map-bt, echo = FALSE, fig.cap = "Bull-trout per-segment mapping_code across the Parsnip River Watershed Group, link's bcfishpass configuration. Stream colours come straight from the bcfishpass symbology registry bundled in gq, so they match a bcfishpass QGIS project exactly. Context: lakes/rivers/manmade waterbodies (light blue), provincial parks (green), First Nations reserves (grey polygon + black diamond + label), resource roads (grey), railways (black dashed); the heavy black line is the watershed-group boundary."}
bt <- streams[!is.na(streams$mapping_code_bt) & nzchar(streams$mapping_code_bt), ]
bt$col <- cls$values[bt$mapping_code_bt]
bt$col[is.na(bt$col)] <- "#999999"

op <- par(mar = c(2, 1, 4, 8))
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, axes = FALSE,
     main = "PARS — bull trout mapping_code\n(link bcfishpass config)")
draw_context_full()
plot(sf::st_geometry(bt), col = bt$col, lwd = 0.6, add = TRUE)
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, add = TRUE)
label_reserves(reserves)
legend_mapping_code(bt$mapping_code_bt)
par(op)
```

## Arctic grayling — a link extension

bcfishpass does not model Arctic grayling, so there is nothing to compare
against; this is net-new output. `link`'s `default` configuration carries `GR`
in its species dimensions, and the same six-phase pipeline that produced the
bull-trout parity above produces a per-segment `mapping_code` for grayling.
The map below is rendered with the **same** gq symbology registry — the token
vocabulary (`ACCESS`/`SPAWN`/`REAR` × `NONE`/`MODELLED`/`ASSESSED`/`DAM`/…) is
species-agnostic, so one colour lookup styles every species consistently.

```{r map-gr, echo = FALSE, fig.cap = "Arctic grayling per-segment mapping_code across the Parsnip River Watershed Group, link's default configuration — a species bcfishpass does not model. Same gq registry symbology and context layers as the bull-trout map, so the two are directly comparable. The grayling network is smaller than bull trout's (19,233 vs 31,932 classified segments), but every grayling segment is net-new output relative to bcfishpass, and 1,764 of them carry no bull-trout classification at all."}
gr <- streams[!is.na(streams$mapping_code_gr) & nzchar(streams$mapping_code_gr), ]
gr$col <- cls$values[gr$mapping_code_gr]
gr$col[is.na(gr$col)] <- "#999999"

op <- par(mar = c(2, 1, 4, 8))
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, axes = FALSE,
     main = "PARS — Arctic grayling mapping_code\n(link default config)")
draw_context_full()
plot(sf::st_geometry(gr), col = gr$col, lwd = 0.6, add = TRUE)
plot(sf::st_geometry(aoi), col = NA, border = "black", lwd = 1.5, add = TRUE)
label_reserves(reserves)
legend_mapping_code(gr$mapping_code_gr)
par(op)
```

## Maps — detail comparison

The full-watershed views compress a lot of network. Cropping to a sub-reach
puts bull trout and grayling side by side at full resolution. Grayling's
modelled network is the smaller of the two overall, but 1,764 segments carry a
grayling classification with no bull-trout classification — the reaches where
the extension is doing genuinely new work.

```{r map-detail, echo = FALSE, fig.width = 7, fig.height = 4.5, fig.cap = "South-east corner of the Parsnip River Watershed Group at full resolution — the headwaters near the continental divide: bull trout (left, link bcfishpass config) and Arctic grayling (right, link default config), same extent, same gq symbology. Grey background streams are the full modelled network, so the coloured overlay shows where each species' classification reaches. Context: waterbodies (light blue), parks (green), reserves (grey + diamond), roads (grey), railways (black dashed), named streams (italic blue labels)."}
e <- sf::st_bbox(aoi)
inset_bbox <- sf::st_bbox(c(
  xmin = unname(e["xmin"] + (e["xmax"] - e["xmin"]) * 0.55),
  ymin = unname(e["ymin"]),
  xmax = unname(e["xmax"]),
  ymax = unname(e["ymin"] + (e["ymax"] - e["ymin"]) * 0.45)
), crs = sf::st_crs(aoi))

crop_sf <- function(x) {
  if (is.null(x) || nrow(x) == 0L) return(NULL)
  out <- suppressWarnings(sf::st_crop(x, inset_bbox))
  if (nrow(out) == 0L) NULL else out
}

frame <- sf::st_as_sfc(inset_bbox)
all_d   <- crop_sf(streams)
bt_d    <- crop_sf(bt)
gr_d    <- crop_sf(gr)
wb_d    <- crop_sf(waterbodies)
aoi_d   <- crop_sf(aoi)
roads_d <- crop_sf(roads)
rail_d  <- crop_sf(railways)
res_d   <- crop_sf(reserves)
parks_d <- crop_sf(parks)
ns_d    <- crop_sf(named_streams)

panel <- function(seg, title) {
  plot(sf::st_geometry(frame), col = NA, border = NA, axes = FALSE,
       main = title)
  if (!is.null(wb_d)) {
    plot(sf::st_geometry(wb_d), col = "#a3cdb988", border = "#2171B5",
         lwd = 0.4, add = TRUE)
  }
  if (!is.null(all_d)) {
    plot(sf::st_geometry(all_d), col = "#cccccc", lwd = 0.4, add = TRUE)
  }
  if (!is.null(roads_d)) {
    plot(sf::st_geometry(roads_d), col = "#666666aa", lwd = 0.4, add = TRUE)
  }
  if (!is.null(rail_d)) {
    plot(sf::st_geometry(rail_d), col = "black", lwd = 1.0, lty = "dashed",
         add = TRUE)
  }
  if (!is.null(parks_d)) {
    plot(sf::st_geometry(parks_d), col = "#639b5f55", border = "#33a02c",
         lwd = 0.8, add = TRUE)
  }
  if (!is.null(res_d)) {
    plot(sf::st_geometry(res_d), col = "#b2b2b2aa", border = "#232323",
         lwd = 0.7, add = TRUE)
    rpts <- suppressWarnings(sf::st_centroid(sf::st_geometry(res_d),
                                             of_largest_polygon = TRUE))
    plot(rpts, pch = 23, bg = "#222222", col = "white", cex = 1.0, add = TRUE)
  }
  if (!is.null(seg) && nrow(seg) > 0L) {
    plot(sf::st_geometry(seg), col = seg$col, lwd = 0.9, add = TRUE)
  }
  if (!is.null(aoi_d)) {
    plot(sf::st_geometry(aoi_d), border = "black", lwd = 1.5, add = TRUE)
  }
  if (!is.null(ns_d)) {
    dedup <- ns_d[!duplicated(ns_d$gnis_name), ]
    npts <- suppressWarnings(sf::st_centroid(sf::st_geometry(dedup)))
    text(sf::st_coordinates(npts), labels = dedup$gnis_name, cex = 0.45,
         font = 3, col = "#0d3a6c")
  }
}

op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
panel(bt_d, "Bull trout")
panel(gr_d, "Arctic grayling")
par(op)
```

## From vignette to report

This vignette is a rehearsal. The same analysis — parity to earn trust, then a
species extension that bcfishpass does not cover — is built to transfer into
the **Fish Passage Peace 2025** report as an appendix, following the path the
`flooded` package's floodplain vignette took into its own reporting template.
The framing there is deliberate and load-bearing: `link` *complements and
extends* the `fwapg` / `bcfishpass` / `bcfishobs` stack, and the report should
say so plainly. The parity number is what licenses the extension — once
`link` reproduces bcfishpass for bull trout, its grayling output stands on the
same modelling, just pointed at a species the upstream stack leaves unmodelled.

The grayling showcase here uses the `default` configuration and is independent
of any work to add grayling to the `bcfishpass`-comparison config — the two
should not be conflated.
