---
title: "Reproducing bcfishpass with link + fresh"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Reproducing bcfishpass with link + fresh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 72
)
library(link)
```

## Why this vignette

[bcfishpass](https://github.com/smnorris/bcfishpass) is the reference
model for freshwater habitat classification and fish passage
prioritization in British Columbia. link + [fresh](https://newgraphenvironment.github.io/fresh/)
together give you a configurable, reproducible R-side pipeline that
can reproduce bcfishpass's habitat classification — and, with a
different config bundle, run the same machinery with different
interpretation choices.

This vignette walks through the `"bcfishpass"` config: load the
bundle, run the pipeline across four watershed groups, inspect the
rollup, and visualize one result. The goal is **bit-identical output
from the same inputs** — parity against bcfishpass reference tables
is an informational comparison, not the correctness bar.

## The three-line entrypoint

Every run of the comparison pipeline follows this shape:

```{r entrypoint, eval = FALSE}
library(link)
library(targets)

# `_targets.R` lives in data-raw/ — run from that directory
# or use targets::tar_config_set(script = "data-raw/_targets.R").
setwd("data-raw")

tar_make()                  # ~8.5 min serial, 4 WSGs
rollup <- tar_read(rollup)  # tibble of per-WSG / per-species diffs
```

`tar_make()` runs `compare_bcfishpass_wsg()` once per watershed group
(ADMS, BULK, BABL, ELKR). Each call exercises the six
`lnk_pipeline_*` phases — setup, load, prepare, break, classify,
connect — then queries the bcfishpass reference tables on a read-only
tunnel DB and returns a ~10-row tibble.

## The targets graph

Inside each per-WSG target, the pipeline looks like this:

```
lnk_config("bcfishpass") ─┬─► comparison_ADMS ─┐
                          ├─► comparison_BULK ─┤
                          ├─► comparison_BABL ─┼─► rollup (34 rows)
                          └─► comparison_ELKR ─┘
```

See `research/bcfishpass_comparison.md` in the repo for:

- the full per-phase pipeline DAG (setup → load → prepare → break →
  classify → connect, with data inputs) rendered as Mermaid
- a colour-coded targets orchestration diagram
- per-species / per-watershed-group parity numbers
- documented gaps and the interpretation decisions that drive them

## The rollup

The rollup binds every per-WSG tibble into one 34-row frame.

```{r load-rollup}
rollup_path <- system.file("extdata", "vignette-data", "rollup.rds",
                            package = "link")
rollup <- readRDS(rollup_path)
rollup
```

Interpretation:

- `link_km` — kilometres our pipeline classified as habitat
  (spawning or rearing) per species × watershed group
- `bcfishpass_km` — kilometres from bcfishpass's reference
  `habitat_linear_*` tables
- `diff_pct` — informational: `(link_km − bcfishpass_km) / bcfishpass_km × 100`

All 34 rows fall within 5% of the bcfishpass reference. Perfect
agreement would be `diff_pct = 0`; modest differences arise from
segmentation boundary choices, rounding in `downstream_route_measure`,
and deliberate interpretation differences baked into the bcfishpass
config's rules YAML.

```{r rollup-summary}
summary(rollup$diff_pct)
cat("Within 5%: ", all(abs(rollup$diff_pct[!is.na(rollup$diff_pct)]) < 5))
```

## What a result looks like

Numbers on a table give you the aggregate — a map tells you where the
habitat is. Below is BULK's chinook (CH) habitat, filtered to segments
classified as spawning, rearing, or both.

```{r load-bulk-ch, message = FALSE}
bulk_ch_path <- system.file("extdata", "vignette-data", "bulk_ch.rds",
                             package = "link")
bulk_ch <- readRDS(bulk_ch_path)
bulk_ch
```

```{r map-bulk-ch, message = FALSE, warning = FALSE}
if (requireNamespace("mapgl", quietly = TRUE)) {
  mapgl::maplibre(
    bounds = sf::st_bbox(bulk_ch),
    style = mapgl::carto_style("positron")
  ) |>
    mapgl::add_line_layer(
      id = "bulk_ch",
      source = bulk_ch,
      line_color = mapgl::match_expr(
        "habitat",
        values = c("spawning only",
                    "rearing only",
                    "spawning + rearing"),
        stops  = c("#e31a1c", "#1f78b4", "#6a3d9a"),
        default = "#999999"
      ),
      line_width = 2
    ) |>
    mapgl::add_legend(
      "BULK · chinook habitat",
      values = c("spawning only",
                  "rearing only",
                  "spawning + rearing"),
      colors = c("#e31a1c", "#1f78b4", "#6a3d9a"),
      type = "categorical"
    )
} else {
  message("Install `mapgl` (pak::pak('mapgl')) to render this map.")
}
```

Rearing dominates the network (CH juveniles use mainstems
extensively); spawning is more focused — low gradient, appropriate
channel width. Where the two overlap, the segment supports both life
stages.

## Reproducibility

The pipeline is **deterministic**. Two `tar_make()` invocations on
the same fwapg + bcfishobs state produce bit-identical rollups. If
you observe drift between runs with the same inputs, that is a bug.

If input data shifts (a fresh `channel_width` sync, new observations
loaded into `bcfishobs`, a bcfishpass reference refresh), outputs
will correctly differ. The lineage tooling tracked in
[link#40](https://github.com/NewGraphEnvironment/link/issues/40) will
make every output traceable to exactly the inputs that produced it.

## Where newgraph's own interpretation lives

This vignette runs the `"bcfishpass"` config — reproduces bcfishpass
behavior exactly, for validation. A companion `"default"` config
bundle is scaffolded at `inst/extdata/configs/default/` and will
carry newgraph's own interpretation choices (intermittent streams
included for rearing, sharper spawn gradient minimums, expanded lake
rearing rules). The biology behind those departures is tracked in
[link#19](https://github.com/NewGraphEnvironment/link/issues/19),
[#20](https://github.com/NewGraphEnvironment/link/issues/20),
[#21](https://github.com/NewGraphEnvironment/link/issues/21); a
second vignette will compare the two configs side by side once those
land.

## Further reading

- `research/bcfishpass_comparison.md` — per-phase pipeline DAG,
  remaining gaps, break sources, key fixes
- `?lnk_pipeline_setup` etc. — reference for each phase helper
- `?lnk_config` — config bundle structure
- `data-raw/_targets.R` — pipeline definition
- `data-raw/compare_bcfishpass_wsg.R` — per-AOI target function
