---
title: "Reproducing bcfishpass with link + fresh"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Reproducing bcfishpass with link + fresh}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 8,
  fig.height = 6,
  dpi = 72
)
library(link)
```

## Why this vignette

[bcfishpass](https://github.com/smnorris/bcfishpass) is the reference
model for freshwater habitat classification and fish passage
prioritization in British Columbia. link + [fresh](https://newgraphenvironment.github.io/fresh/)
together give you a configurable, reproducible R-side pipeline that
can reproduce bcfishpass's habitat classification — and, with a
different config bundle, run the same machinery with different
interpretation choices.

This vignette walks through the `"bcfishpass"` config: the story of
what the pipeline does to an FWA stream network, how to run it, what
comes out the other side, and where the bundled config deliberately
departs from bcfishpass.

## How bcfishpass works under the hood

The question the pipeline answers is, for every segment of the
Freshwater Atlas stream network: **can this species reach this
segment, and is it suitable as spawning or rearing habitat?** The
answer for one segment × one species is a pair of booleans. The work
is producing those booleans for ~50,000 segments per watershed group
and a dozen species.

```
FWA streams (raw)
    │
    │   thresholds detect gradient barriers @ 15 / 20 / 25 / 30 %
    ▼
gradient barriers ─── falls ─── user-identified definite barriers
    │
    │   observations override natural barriers upstream of evidence
    ▼
access model per species  (which reaches are reachable for BT, CH, CO, ...)
    │
    │   break positions = fish observations + minimal gradient barriers
    │                     + expert-confirmed habitat endpoints + crossings
    ▼
segmented streams (every segment ends where a classification decision changes)
    │
    │   per-species rules from the rules YAML:
    │     edge type • waterbody type • channel width • gradient
    ▼
classify  (spawning ?  rearing ? per species per segment)
    │
    │   connectivity — rearing requires reach from spawning;
    │                   SK spawning requires a reachable rearing lake
    ▼
streams_habitat  (per-species spawning / rearing booleans on every segment)
```

### Where breaks go and why

A break is a point on the stream where one segment ends and the next
begins. A segment is a unit of classification — one segment, one
answer per species. So breaks go wherever the answer could change:

- **Fish observations.** A confirmed observation upstream of a
  gradient barrier means the barrier can't actually be a barrier for
  that species — the pipeline flips the access label on that reach
  from "blocked" to "passable."
- **Minimal gradient barriers.** On a flow path with multiple
  gradient barriers, only the downstream-most one matters for access
  — everything upstream of it is already blocked by it. The pipeline
  reduces to this minimal set per species-class so the segmentation
  doesn't fragment reaches that were going to end up in the same
  access state anyway.
- **Expert-confirmed habitat endpoints.** Where a human review has
  marked a specific reach boundary as habitat / non-habitat, the
  pipeline breaks on both ends so the marked reach is its own
  segment.
- **Crossings.** Road and rail crossings are potential barriers, and
  their `barrier_status` (PASSABLE / POTENTIAL / BARRIER) is applied
  per-segment.

### Where classification comes from

Once the network is segmented, each segment is checked against the
per-species rules in the config's rules YAML:

```r
rules_path <- system.file("extdata", "configs", "bcfishpass",
                           "rules.yaml", package = "link")
cat(readLines(rules_path, n = 30), sep = "\n")
```

Each top-level key is a species. `spawn:` and `rear:` are lists of
alternative match conditions — a segment is classified as spawning
(or rearing) if any one condition matches. Conditions combine
`edge_types_explicit` (1000 / 1100 streams, 2000 / 2300 rivers),
`waterbody_type` (`R` river polygon, `L` lake), `channel_width`
ranges, and gradient bounds. A species can also flag `rear_lake`,
`rear_lake_only`, or `rear_no_fw` to special-case its rearing
strategy. The `dimensions.csv` alongside the YAML is the
human-editable source — `lnk_rules_build()` regenerates the YAML
from it.

### Where link + fresh deliberately depart

bcfishpass applies a stream-order bypass on the rearing
channel-width minimum for BT, CH, CO, ST, and WCT: a first-order
stream whose parent is order ≥ 5 is allowed to count as rearing
even if its measured channel width is below the per-species minimum.
The biology is that small tributaries of major rivers can still
support juvenile rearing despite narrow measured widths — a
reasonable move.

The link pipeline does not currently apply this rule. The bypass is
awkward to compose with the clustering step — applying it before
clustering inflates rearing along tributaries that have no
connectivity to spawning, and applying it after clustering adds
segments without the connectivity check. We're working on a
biologically cleaner alternative that restricts the bypass to direct
first-order children of large-order streams within a distance cap
from the tributary mouth, which we'll wire into a future config
variant. In the current bundled `"bcfishpass"` config the practical
consequence is a 68 km shortfall in BT rearing on BULK — the −2.2 %
entry you'll see in the rollup below. Every other species / WSG
combination agrees with the reference within about 5 %.

## The three-line entrypoint

```{r entrypoint, eval = FALSE}
library(link)
library(targets)

# `_targets.R` lives in data-raw/ — run from that directory
# or use targets::tar_config_set(script = "data-raw/_targets.R").
setwd("data-raw")

tar_make()                  # ~8.5 min serial, 4 WSGs
rollup <- tar_read(rollup)  # tibble of per-WSG / per-species diffs
```

`tar_make()` runs `compare_bcfishpass_wsg()` once per watershed group
(ADMS, BULK, BABL, ELKR). Each call exercises the six
`lnk_pipeline_*` phases — setup, load, prepare, break, classify,
connect — then queries the bcfishpass reference tables on a read-only
tunnel DB and returns a small tibble.

## The targets graph

```
lnk_config("bcfishpass") ─┬─► comparison_ADMS ─┐
                          ├─► comparison_BULK ─┤
                          ├─► comparison_BABL ─┼─► rollup (34 rows)
                          └─► comparison_ELKR ─┘
```

See `research/bcfishpass_comparison.md` in the repo for the per-phase
pipeline DAG rendered as Mermaid plus a colour-coded targets
orchestration view.

## The rollup

```{r load-rollup}
rollup_path <- system.file("extdata", "vignette-data", "rollup.rds",
                            package = "link")
rollup <- readRDS(rollup_path)
```

```{r rollup-summary}
summary(rollup$diff_pct)
cat("Within 5%: ",
    all(abs(rollup$diff_pct[!is.na(rollup$diff_pct)]) < 5))
```

- `link_km` — kilometres our pipeline classified as habitat
  (spawning or rearing) per species × watershed group
- `bcfishpass_km` — kilometres from bcfishpass's reference
  `habitat_linear_*` tables
- `diff_pct` — `(link_km − bcfishpass_km) / bcfishpass_km × 100`

Modest differences arise from segmentation boundary choices, rounding
in `downstream_route_measure`, the stream-order bypass discussed
above, and deliberate interpretation differences baked into the
bcfishpass rules YAML.

## What a result looks like

Numbers on a table give you the aggregate — a map tells you where the
habitat is. Below is BULK chinook habitat, rendered from the same
`fresh.streams_habitat` table the rollup aggregates.

```{r map-bulk-ch, message = FALSE, warning = FALSE}
bulk_ch_path <- system.file("extdata", "vignette-data", "bulk_ch.rds",
                             package = "link")
bulk_ch <- readRDS(bulk_ch_path)

if (requireNamespace("mapgl", quietly = TRUE)) {
  mapgl::maplibre(
    bounds = sf::st_bbox(bulk_ch),
    style = mapgl::carto_style("positron")
  ) |>
    mapgl::add_line_layer(
      id = "bulk_ch",
      source = bulk_ch,
      line_color = mapgl::match_expr(
        "habitat",
        values = c("spawning only",
                    "rearing only",
                    "spawning + rearing"),
        stops  = c("#e31a1c", "#1f78b4", "#6a3d9a"),
        default = "#999999"
      ),
      line_width = 2
    ) |>
    mapgl::add_legend(
      "BULK · chinook habitat",
      values = c("spawning only",
                  "rearing only",
                  "spawning + rearing"),
      colors = c("#e31a1c", "#1f78b4", "#6a3d9a"),
      type = "categorical"
    )
} else {
  message("Install `mapgl` (pak::pak('mapgl')) to render this map.")
}
```

Rearing dominates the network — CH juveniles use mainstems
extensively. Spawning is more focused — low gradient, appropriate
channel width. Where the two overlap the segment supports both life
stages.

## Reproducibility

The pipeline is deterministic. Two `tar_make()` invocations on the
same fwapg + bcfishobs state produce bit-identical rollups. If you
observe drift between runs with the same inputs, that's a bug. When
input data shifts — a fresh `channel_width` sync, new observations
loaded into `bcfishobs`, a bcfishpass reference refresh — outputs
will correctly differ.

## Further reading

- `research/bcfishpass_comparison.md` — per-phase pipeline DAG,
  documented gaps, break sources, key fixes
- `?lnk_pipeline_setup` etc. — reference for each phase helper
- `?lnk_config` — config bundle structure
- `inst/extdata/configs/bcfishpass/rules.yaml` — per-species
  classification rules for this variant
- `inst/extdata/configs/bcfishpass/dimensions.csv` — human-editable
  source of the rules YAML
- `data-raw/_targets.R` — pipeline definition
- `data-raw/compare_bcfishpass_wsg.R` — per-AOI target function
