---
title: "Scoring crossings for fish passage prioritization"
output: bookdown::html_vignette2
vignette: >
  %\VignetteIndexEntry{Scoring crossings for fish passage prioritization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(link)
```

This vignette walks through the crossing interpretation pipeline using
the Morice watershed group (MORR) as an example. The same steps apply to
any watershed group in the province.

Data generated by `data-raw/vignette_morr.R`. Sources: crossings from
[fresh](https://github.com/NewGraphEnvironment/fresh) CSV, PSCIS
assessments from BC Data Catalogue via
[bcdata](https://bcgov.github.io/bcdata/), override CSVs from
[bcfishpass](https://github.com/smnorris/bcfishpass). No database
required.

## How link fits with fresh

[fresh](https://newgraphenvironment.github.io/fresh/) is a generic
stream network engine. It segments streams at break points, classifies
habitat by species thresholds, and rolls up results. But fresh doesn't
know about crossings, severity, or field assessments --- that's link's
job.

```
link (interpret, score crossings) --> break source spec --> fresh (segment, classify habitat)
```

link reads raw crossing data, applies field corrections, scores severity,
and produces a `break_sources` spec that plugs directly into
`frs_habitat()`. Confirmed barriers break the network geometry. Unassessed
crossings are features registered on the network for downstream indexing
(see fresh [#92](https://github.com/NewGraphEnvironment/fresh/issues/92),
[#93](https://github.com/NewGraphEnvironment/fresh/issues/93)).

## The problem with binary classification

Provincial crossing assessments produce BARRIER or PASSABLE. This tells
you nothing about how bad a barrier is or what habitat it blocks.

```{r raw-crossings}
morr_raw <- readRDS(system.file(
  "testdata", "morr_crossings_raw.rds", package = "link"))

nrow(morr_raw)
table(morr_raw$crossing_source)
table(morr_raw$barrier_status)
```

MORR has `r nrow(morr_raw)` crossings ---
`r sum(morr_raw$crossing_source == "MODELLED CROSSINGS")` modelled,
`r sum(morr_raw$crossing_source == "PSCIS")` with PSCIS field
assessments. Most modelled crossings have no field data.

## Override corrections

bcfishpass maintains correction CSVs --- tens of thousands of
hand-reviewed crossings accumulated across field seasons and imagery
review.

```{r overrides}
morr_fixes <- readRDS(system.file(
  "testdata", "morr_modelled_fixes.rds", package = "link"))

nrow(morr_fixes)
table(morr_fixes$structure)
```

For MORR: `r nrow(morr_fixes)` modelled crossing fixes. Most reclassify
crossings as empty (no structure found on imagery) or open-bottom
structures (passable).

```{r overrides-sample}
head(morr_fixes[, c("modelled_crossing_id", "structure",
                     "reviewer_name", "source")], 8)
```

The link override pipeline loads, validates, and applies these
corrections:

```{r overrides-pipeline, eval = FALSE}
conn <- lnk_db_conn()

# Load --- validates CSV structure before writing to DB
lnk_override_load(conn,
  csv  = "user_modelled_crossing_fixes_morr.csv",
  to   = "working.morr_modelled_fixes",
  cols_id = "modelled_crossing_id",
  cols_required = c("structure"),
  cols_provenance = c("reviewer_name", "review_date", "source"))

# Validate --- find orphans and duplicates
lnk_override_validate(conn,
  overrides = "working.morr_modelled_fixes",
  crossings = "working.morr_crossings")

# Apply --- auto-detects which columns to update
lnk_override_apply(conn,
  crossings = "working.morr_crossings",
  overrides = "working.morr_modelled_fixes")
```

## PSCIS matching

PSCIS assessments have field measurements --- outlet drop, culvert slope,
length, channel width. The crossings table from fresh has network position.
Matching links the measurements to the network so severity scoring uses
real data.

```{r pscis}
morr_pscis <- readRDS(system.file(
  "testdata", "morr_pscis.rds", package = "link"))
morr_xref <- readRDS(system.file(
  "testdata", "morr_xref.rds", package = "link"))

nrow(morr_pscis)
nrow(morr_xref)
```

The `r nrow(morr_xref)` xref corrections are hand-curated GPS error fixes
from field work. These take priority over spatial matching.

```{r pscis-match, eval = FALSE}
lnk_match_pscis(conn,
  crossings = "working.morr_crossings",
  pscis     = "working.morr_pscis",
  xref_csv  = "pscis_modelledcrossings_streams_xref_morr.csv",
  to        = "working.morr_matched")
```

## PSCIS barrier status overrides

```{r pscis-status}
morr_pscis_fixes <- readRDS(system.file(
  "testdata", "morr_pscis_status_fixes.rds", package = "link"))

if (nrow(morr_pscis_fixes) > 0) {
  morr_pscis_fixes[, c("stream_crossing_id",
                        "user_barrier_status", "notes")]
}
```

One MORR crossing has a barrier status override --- confirmed PASSABLE by
imagery review.

## Severity scoring

Default thresholds classify crossings by biological impact:

```{r thresholds}
th <- lnk_thresholds()
```

| Severity | Criteria | Interpretation |
|----------|----------|----------------|
| High | outlet_drop >= `r th$high$outlet_drop`m OR slope x length >= `r th$high$slope_length` | Impassable at most flows |
| Moderate | outlet_drop >= `r th$moderate$outlet_drop`m OR slope x length >= `r th$moderate$slope_length` | Flow-dependent |
| Low | everything else | Likely passable |

```{r score, eval = FALSE}
lnk_score_severity(conn, "working.morr_crossings")
```

```{r severity-dist}
morr_dist <- readRDS(system.file(
  "testdata", "morr_severity_dist.rds", package = "link"))
morr_dist
```

```{r severity-detail}
morr_scored <- readRDS(system.file(
  "testdata", "morr_pscis_scored.rds", package = "link"))

# The reveal: crossings with PSCIS measurements scored by severity
scored_with_data <- morr_scored[!is.na(morr_scored$severity), ]
head(scored_with_data[order(-scored_with_data$outlet_drop),
  c("stream_crossing_id", "barrier_result_code",
    "outlet_drop", "culvert_slope",
    "severity")], 10)
```

The binary `barrier_result_code` treats these crossings identically. The
`severity` column reveals the gradient --- from impassable drops to likely
passable culverts.

## Handing off to fresh

`lnk_break_source()` produces the spec that plugs directly into
[fresh](https://newgraphenvironment.github.io/fresh/)'s
`frs_habitat()`. Link's severity labels translate to fresh's access
labels via `label_map`:

- `high` severity --> `"blocked"` (confirmed barrier, breaks geometry)
- `moderate` severity --> `"potential"` (unconfirmed, may break geometry)

```{r break-source}
morr_spec <- readRDS(system.file(
  "testdata", "morr_break_source.rds", package = "link"))
str(morr_spec)
```

```{r break-source-use, eval = FALSE}
src <- lnk_break_source(conn, "working.morr_crossings")

# The handoff --- fresh takes it from here
fresh::frs_habitat(conn, "MORR", break_sources = list(src))

# After fresh classifies habitat, link can roll up per crossing
lnk_habitat_upstream(conn, "working.morr_crossings",
  "fresh.streams_habitat")
```

## Scaling to any watershed group

The same steps work for any `watershed_group_code`. The override CSVs
have `watershed_group_code` columns for filtering. The crossings CSV from
fresh covers the entire province.

```{r province, eval = FALSE}
crossings <- read.csv(system.file("extdata", "crossings.csv",
                                   package = "fresh"))
wsgs <- unique(crossings$watershed_group_code)

for (wsg in wsgs) {
  # 1. Filter crossings to WSG
  # 2. Load and apply overrides (filtered to WSG)
  # 3. Fetch PSCIS from bcdata, match with lnk_match_pscis()
  # 4. lnk_score_severity()
  # 5. lnk_break_source() -> frs_habitat()
}
```
